The Spine of American Legal Practice

### 2016-06-28 article/article.Rmd
Revise part 4

## Part 4

Alerted by these clustering patterns to the significance of creditors' remedies in the developing West and re-developing South, it was not difficult to find evidence of westerners and southerners commenting on these features of the code or its connections to New York capital. Twelve of the states and territories that copied the Field Code mostly closely did so during the Civil War and Reconstruction era---four states in the former Confederacy and eight jurisdictions in the Far West.[^49] As with other areas of postbellum historiography, it turns out one may learn a lot by holding the postbellum American South and American West together.

In the last decade, scholars of Reconstruction have broadened the scope of their study to include both the South and the West as two sites in one "Greater Reconstruction." These studies have illustrated the ways that military conquest, rapid industrialization, and the resettlement and education of ethnic minorities developed similarly in each region, guided by economic elites in New York and political administrators in Washington.[^50] In tracing the legal aspects of this Greater Reconstruction, scholars have focused almost entirely on constitutional rights of citizenship and civil equality or the expansion of federal power.[^51] While the 1860s and 1870s were of course a transformative period in the history of civil rights and the creation of a national state, they were also the decades in which many local legal institutions and practices were transformed not by federal power but by state codification. Naomi Lamoreaux and John Joseph Wallis have recently argued that in the creation of a modern American economy, "the federal government played *no* role in this process until the Civil War, and even then it played only a bit part." The history of the Field Code's migration helps to substantiate this claim. While Lamoreaux and Wallis focus on the development of banking, transportation, and incorporation at the state level, it was the Procedure Code that provided the structure of civil remedies that protected these state institutions. And procedure codes were, of course, the creatures of state governance.[^52]

With the universal complaints that time was too short or commissioners too democratically irresponsible, one might expect the code to have faced more competition from other workable statutes. Southern states that drew more on the civilian code tradition such as Louisiana and Alabama offered other options for states desiring to reform common law practice.[^53] And if Euro-Americans believed law had to be imported into the West, Colorado immigrants proved that the common law of Illinois was just as amenable to statutory importation as New York law.[^54] But what codifiers saw when they looked at New York, more so than Louisiana, Alabama, or Illinois, was the Empire State of commercial capital. And the putative fears, demands, and desires of a personified Capital continually wielded promises---and threats---in the debates over procedural codification. The early Mormon settlers of Utah persistently avoided the mining frenzy as well as codification. By 1870, however, the territory's governor directed the legislature's attention to the recent Code of Nevada, a code "for a people whose interests in many respects are similar to our own." Of course, standing behind the Nevada Code was "the State of New York---a State which is an empire in itself and whose commercial transactions are far greater than those of any other State in the Union." By copying its code, Utah too could be "rewarded by equal advantages."[^55] Code proponents in Colorado similarly parried arguments that a code would be a "foreign" imposition by pointing to the fact that it has been "adopted twenty-nine years ago by the Empire state of the Union" and they promised that the code of the nation's uniquely commercial empire could bring wealth in its wake.[^56]

@@

Receiving innumerable letters complaining that under the code, "no one will be benefited, except perhaps some Northern Capitalists," a North Carolina commissioner undertook an anonymous defense of the new Code in the *Weekly Standard*. He encouraged the bar to accommodate themselves to change, for "the New York system ... bids fair to become national." Putatively offering an overview of the code, the articles were almost entirely about credit. "How can we create credit? By punctuality," the commissioner wrote. "And how create punctuality? by law, and by law alone. Let the law enforce punctuality; let the people of North Carolina learn that the great law of business is, that 'time is of the essence of the contract,' and incur no debt that they do not expect to meet at maturity." Under the more certain and speedy remedies of the code, "we may expect that the secret hoards of the frugal among ourselves will be offered to loan, and even that the vaults of the banks of New-York ... will be open to our industry."[^59]

As in postbellum North Carolina, establishing a flow of credit through the remedial system became a leading priority of western lawyers. While the so-called new western history has shed significant light on neglected topics of Indian dispossession and environmental management, it has often done so by leaving out of view matters of political economy, a staple of the old western history. As one work in that older tradition argued, "Debt collection was the central part of law practice for the \[western\] bar and remained a key part of private practice throughout the century."[^60] On that understanding, one western lawyer succinctly summarized the difference between the code and the common law as "whether a merchant had better try to collect a \$500 note or burn it up." Tiring of all the focus on creditors' remedies, one legislator observed that he "never knew one of these professionals who undertook to write up the beauties of the New York code, ... that he did not also break out somewhere with 'take for instance the case of an action on a promissory note,' as though the collection of notes was about all there could be any law needed for."[^61]

@@

Thus by the end of Reconstruction, New York's domestic empire of capital and creditors' remedies bore a remarkable resemblance to the international empire administered by England. Both jurisdictions, while reforming the practice of law, largely rejected codification within their own borders but encouraged it among their economic dependents. The English commissioned a full codification of law for India and colonies in Singapore, while further codifications produced by Field in New York covering civil and penal law were adopted in California and other western jurisdictions.[^64] In both places, leading arguments against codification carried a civilizational logic of empire: Advanced societies could not codify their law, for to do so would be to freeze the progress of legal science. What appeared to some to be a hopeless mass of confusion was to common law defenders the sign of true legal sophistication. Science was, after all, sophisticated.[^65] The Throop Code came in for censure, both in its first year and in following decades, precisely for trying to capture all the sophistication of the New York legal system within an unwieldy 3,300 rules.[^66] Codification, however, could help developing societies along law's frontier take a progressive leap forward. As India's chief codifier Thomas Macaulay explained, codification "cannot be well performed in an age of barbarism," but also "cannot without great difficulty be performed in an age of freedom." As India balanced between the two, however, "it is the work which especially belongs to a government like that of India---to an enlightened and paternal despotism."[^67]

In America, Macaulay's tool of enlightened despotism spread with the anxiety that capital from the nation's economic center would remain scarce without a Code of remedies that, if not in fact the law of New York, was at least prescribed by New York lawyers and their corporate clients. In the two most populous and commercially advanced western states, Texas and Illinois, the need for New York capital failed to move state legislators to adopt the code at the expense of popular sovereignty (although there were concerted efforts in both jurisdictions).[^68] Lacking the self-sufficiency of those two jurisdictions, the other states of the Greater Reconstruction in America adopted a foreign Code, but lawyers, legislators, and their supporters claimed the endorsement of popular sovereignty in doing so. Even in North Carolina, whose Democratic newspapers daily called for the repeal "this child of the carpet baggers," Republican editors proclaimed that "the movement" towards procedural codification "comes from the people, from the instinctive logic by which an unprejudiced mind grasps the advantages of the system."[^69]

@@

[^54]: Illinois was a fiercely traditional anti-code state. In later decades, lawyers jokingly called it "the Yellowstone Park of common law pleading." See Charles E. Clark, "The New Illinois Civil Practice Act," 1 *University of Chicago Law Review* 209 (1933). Nevertheless, even Illinois had substantial legislation organizing the courts and prescribing certain common law processes for obtaining civil remedies, and Colorado adapted the bulk of this legislation when it organized as a territory, a full seventeen years before its legislature considered adopting the Field Code.

[^55]: Journal of the Assembly of the Territory of Utah (1870), 15.

>>>>

## 4. Procedure codes and capitalism

The near uniformity of creditors' remedies in Field Code states, as demonstrated by clustering, points out the close link between the rise of civil procedure and capitalism. Westerners and southerners commented on these features of the code or its connections to New York capital. Twelve of the states and territories that copied the Field Code most closely did so during the Civil War and Reconstruction era---four states in the former Confederacy and eight jurisdictions in the Far West.[^49] As with other areas of postbellum history, it turns out one may learn a lot by holding the postbellum American South and American West together.

In the last decade, scholars of Reconstruction have broadened the scope of their study to include both the South and the West as two sites in one "Greater Reconstruction." These studies have illustrated the ways that military conquest, rapid industrialization, and the resettlement and education of ethnic minorities developed similarly in each region, guided by economic elites in New York and political administrators in Washington.[^50] In tracing the legal aspects of this Greater Reconstruction, scholars have focused almost entirely on constitutional rights of citizenship and civil equality or the expansion of federal power.[^51] While the 1860s and 1870s were of course a transformative period in the history of civil rights and the creation of a national state, they were also the decades in which many local legal institutions and practices were transformed not by federal power but by state codification. Naomi Lamoreaux and John Joseph Wallis have recently argued that in the creation of a modern American economy, "the federal government played *no* role in this process until the Civil War, and even then it played only a bit part." The history of the Field Code's migration helps to substantiate this claim. While Lamoreaux and Wallis focus on the development of banking, transportation, and incorporation at the state level, it was the Procedure Code that provided the structure of civil remedies that protected these state institutions. And procedure codes were the creatures of state governance.[^52]

The states that adopted the Field Code were never forced to, and they had other options available to them. Southern states that drew more on the civilian code tradition such as Louisiana and Alabama offered other options for states desiring to reform common law practice.[^53] Illinois long retained the common law---in later decades lawyers called it "the Yellowstone Park of common law pleading." When Colorado was a territory it imported the Illinois common law to the West via statute, a full fourteen years before its legislature considered the Field Code.  [^54]

But what codifiers saw when they looked at New York, more so than at Louisiana, Alabama, or Illinois, was the Empire State of commercial capital. The fears, demands, and desires of a personified Capital continually wielded promises---and threats---in the debates over procedural codification. The early Mormon settlers of Utah persistently avoided the mining frenzy as well as codification. By 1870, however, the territory's governor directed the legislature's attention to the recent Code of Nevada, a code "for a people whose interests in many respects are similar to our own." Of course, standing behind the Nevada Code was "the State of New York---a State which is an empire in itself and whose commercial transactions are far greater than those of any other State in the Union." By copying its code, Utah too could be "rewarded by equal advantages."[^55] Code proponents in Colorado similarly parried arguments that a code would be a "foreign" imposition by pointing to the fact that it has been "adopted twenty-nine years ago by the Empire state of the Union" and they promised that the code of the nation's uniquely commercial empire could bring wealth in its wake.[^56]

@@

Receiving innumerable letters complaining that under the code, "no one will be benefited, except perhaps some Northern Capitalists," a North Carolina commissioner undertook an anonymous defense of the new Code in the *Weekly Standard*. He encouraged the bar to accommodate themselves to change, for "the New York system ... bids fair to become national." Supposedly offering an overview of the code, the articles were almost entirely about credit. "How can we create credit? By punctuality," the commissioner wrote. "And how create punctuality? by law, and by law alone. Let the law enforce punctuality; let the people of North Carolina learn that the great law of business is, that 'time is of the essence of the contract,' and incur no debt that they do not expect to meet at maturity." Under the more certain and speedy remedies of the code, "we may expect that the secret hoards of the frugal among ourselves will be offered to loan, and even that the vaults of the banks of New-York ... will be open to our industry."[^59]

As in postbellum North Carolina, establishing a flow of credit through the remedial system became a leading priority of western lawyers. While the new western history has shed significant light on neglected topics of Indian dispossession and environmental management, it has often done so by leaving out of view matters of political economy, a staple of the old western history. As one work in that older tradition argued, "Debt collection was the central part of law practice for the \[western\] bar and remained a key part of private practice throughout the century."[^60] On that understanding, one western lawyer succinctly summarized the difference between the code and the common law as "whether a merchant had better try to collect a \$500 note or burn it up." Tiring of all the focus on creditors' remedies, one legislator observed that he "never knew one of these professionals who undertook to write up the beauties of the New York code, ... that he did not also break out somewhere with 'take for instance the case of an action on a promissory note,' as though the collection of notes was about all there could be any law needed for."[^61]

@@

Thus by the end of Reconstruction, New York's domestic empire of capital and creditors' remedies bore a remarkable resemblance to the international empire administered by England. Both jurisdictions, while reforming the practice of law, largely rejected codification within their own borders but encouraged it among their economic dependents. The English commissioned a full codification of law for India and colonies in Singapore, while further codifications produced by Field in New York covering civil and penal law were adopted in California and other western jurisdictions.[^64] In both places, leading arguments against codification carried a civilizational logic of empire: advanced societies could not codify their law, for to do so would be to freeze the progress of legal science. What appeared to some to be a hopeless mass of confusion was to common law defenders the sign of true legal sophistication. Science was, after all, sophisticated.[^65] The Throop Code came in for censure, both in its first year and in following decades, precisely for trying to capture all the sophistication of the New York legal system within an unwieldy 3,300 rules.[^66] Codification, however, could help developing societies along law's frontier take a progressive leap forward. As India's chief codifier Thomas Macaulay explained, codification "cannot be well performed in an age of barbarism," but also "cannot without great difficulty be performed in an age of freedom." As India balanced between the two, however, "it is the work which especially belongs to a government like that of India---to an enlightened and paternal despotism."[^67]

In America, Macaulay's tool of enlightened despotism spread with the anxiety that capital from the nation's economic center would remain scarce without a code of remedies that, if not in fact the law of New York, was at least prescribed by New York lawyers and their corporate clients. In the two most populous and commercially advanced western states, Texas and Illinois, the need for New York capital failed to move state legislators to adopt the code at the expense of popular sovereignty (although there were concerted efforts in both jurisdictions).[^68] Lacking the self-sufficiency of those two jurisdictions, the other states of the Greater Reconstruction in America adopted a foreign code, but lawyers, legislators, and their supporters claimed the endorsement of popular sovereignty in doing so. Even in North Carolina, whose Democratic newspapers daily called for the repeal "this child of the carpet baggers," Republican editors proclaimed that "the movement" towards procedural codification "comes from the people, from the instinctive logic by which an unprejudiced mind grasps the advantages of the system."[^69]

@@

[^54]: See Charles E. Clark, "The New Illinois Civil Practice Act," *University of Chicago Law Review* 1 (1933): 209. Nevertheless, even Illinois had substantial legislation organizing the courts and prescribing certain common law processes for obtaining civil remedies, and Colorado adapted the bulk of this legislation when it organized as a territory, a full seventeen years before its legislature considered adopting the Field Code.

[^55]: Journal of the Assembly of the Territory of Utah (1870), 15. See also Leonard J. Arrington, *Great Basin Kingdom: An Economic History of the Latter-Day Saints, 1830--1900*, new edition (University of Illinois Press, 2004).
